Automated Lensing Learner - I: An automated Strong Lensing Identification Pipeline

### 2017-03-09 ms.tex
Finished subsection on regularization and model performance

%-------------------------------------------------%
\subsection{Machine Learning Algorithm: Logistic Regression}\label{sec:LR}
%-------------------------------------------------%
The problem of detecting gravitational lenses in images falls
under the general category of \em{classification} in machine
learning. In general the task is to find a function that
assigns data points $x$ to one of two or more classes, denoted
by the class label $y$.  This is
equivalent to specifying a decision boundary, or decision
boundaries between the classes in the space of the data points.
(Compare this to
\em{regression} in which the task is to find a function 
$y=f(x)$, where $y$ is a continuous, rather than discrete,
variable.) In our case, we have two classes: lens and
non-lens containing images, and the data points $x$ are the
HOG feature vectors transformed from the training set data.
In this paper we use the Logisitic Regression (LR)
algorithm, for which the decision boundary is a hyperplane .
(The equivalent in the regression setting would be linear
regression.) In LR we determine the optimal hyperplane by
minimizing the objective function
\begin{equation}
\label{log_reg_objective}
L(A,b) = \sum_i \log \left[1 + \exp \left(-y_i (A \cdot x_i + b) \right) \right]
\end{equation}
where $x_i$ is a data point (HOG feature vector), $y_i$ is the
known label for that data point ($1$ for a lens containing
image, $-1$ for a non-lens containing), and $A$ and $b$ are
the parameters of the hyperplane. Eq. \ref{log_reg_objective}
is to be minimized with respect to $A$ and $b$.
Other more complicated machine learning algorithms exist which
do not necessarily produce a linear decision boundary, such as
Support Vector Machines (SVM), Random Forests, and Neural Networks.
training set data \citep{hastie_09}.

The HOG feature vectors in this paper can be very high-dimensional.
When dealing with high-dimensional data, where the number of
dimensions becomes comparable to the number of data points,
overfitting can become an issue. (An example of an extreme case
of overfitting would be if you fit a degree $n$ polynomial to $n$
points.  The polynomial would simply wiggle so that it goes
through every point, and would have no predictive power if you
tried to interpolate or extrapolate.) In machine learning,
overfitting is avoided using different \em{regularization}
techniques. A common choice for logistic regression is to
add a penalty term to Eq. \ref{log_reg_objective}
\begin{equation}\ref{log_reg_regularized}
L_\text{Reg}(A,b,C_\text{LogReg}) = L(A,b) + \frac{1}{2C_\text{LogReg}} \|A\|
\end{equation}
where the norm $\|A\|$ is typically taken as either the
sum of squares of the coefficients ($L_2$ norm) or the sum
of absolute values of the coefficients ($L_1$ norm).  In this
paper we use the former.

The amount of regularization is
controlled by the parameter $C_\text{LogReg}$:
larger values of $C_\text{LogReg}$ correspond to increasing model
complexity.  If $C_\text{LogReg}$ is too large then the model will
overfit, and if it is too small the model will underfit.
To determine whether a model is overfit or underfit, the model is
trained, (i.e. Eq. \ref{log_reg_regularized} is minimized), on a
subset of the data called the \em{training set} and its performance
(goodness of fit) is evaluated both on the the training set and
a set of hold-out data called the \em{test set} that was not used
in constructing the model.
Figure~\ref{fig:regularization} shows the performance of a
model for selected HOG parameters as a function of the regularization
parameter, $C_\text{LogReg}$, for both the training and the test set.
(The performance can be measured by the accuracy, i.e. percent of
images correctly classified, or by some other metric,
such as the area under the
Receiver Operating Characteristic curve (see Section~\ref{sec:ROC}).)

@@

% Logistic regression and regularization
As described in Section~\ref{sec:LR}, LR trains a model with
complexity determined by the regularization parameter coefficient,
$C_\text{LogReg}$.  Increasing model complexity (larger
$C_\text{LogReg}$), will better describe features in the training set.
However, an overly complex model will overfit the training set at the
expense of its performance on any independent test set.  The
regularization parameter ultimately defines the model performance, and
we must perform a parameter search to identify the optimal value for
$C_\text{LogReg}$.

%  Our running over C - score and AUC as characterizations
Figure~\ref{fig:regularization} shows the model performance as a
function of regularization parameter for each data set HST, LSST, and
LSST10.  The solid and dotted lines respectively correspond to the
model performance on the test and training set.

% Training behavior with C
For either characterization of the model, the model performance on the
training set increases and asymptotes with $C_\text{LogReg}$.  With
increasing model complexity, the model better fits the data set it has
been trained on.  This is analogous to fitting a seventh order
polynomial to seven data points, where the fitting function will go
through every point but will not likely predict additional points.

% Test behavior with C
We use the model performance on a test set to determine the best value
for $C_\text{LogReg}$ that will not overfit the training data.  With
increasing model complexity, we are able to better able to capture
features that are generally characteristic of strong lensing systems
with arcs.  However, past a certain $C_\text{LogReg}$, the model
performance on the test set decreases or asymptotes, as it has overfit
the training set.

% Train time with Creg
In red, we show the train time as a function of $C_\text{LogReg}$.
Lower Creg, more regularized, smaller volume to search for the
solution (best fit coefficients), so faster on average.  But, still
has some randomness associated with the path it takes to convergence
(not purely monotonic).

% Scores vs. regularization on larger train/test set with comparable
% grid search in parameterization
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                                        
\begin{figure*}[t]\label{fig:regularization}
\begin{center}
\includegraphics[width=.66\columnwidth]{figures/Creg/HST.pdf}
\includegraphics[width=.66\columnwidth]{figures/Creg/LSST1.pdf}
\includegraphics[width=.66\columnwidth]{figures/Creg/LSST10.pdf}
\caption{AUC of the model with varying LR regularization coefficient
  parameter, $C_{reg}$, used when training the model classifier.  We
  use a subset of the the 10,000 training images to search over the LR
  $C_{reg}$ parameter, training on 8,000 and testing on 1,000.  Each
  panel corresponds to a different mock observation.  From left to
  right: HST, LSST for one year, and LSST for 10 years.  The solid
  blue lines correspond to the AUC of the test set, and the dotted
  blue lines to the AUC of the training set.  To avoid overfitting, we
  choose the smallest parameter for which the AUC of the test set is
  maximal: 5000, 10, and 5000, respectively.  In thin red solid lines,
  we show the train time of the model, which roughly increases in a
  log-log scaling with logistic regression coefficient parameter.}
\end{center}
\end{figure*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%-------------------------------------------------%
\subsection{Effects of Dataset Size on Model Performance}\label{sec:datasize}
%-------------------------------------------------% 
%-------------------------------------------------% 
\subsubsection{Effects of Training Set Size}\label{sec:trainsetsize}
%-------------------------------------------------% 
In this section, we show the effects of training set size on model
performance.  Figure~\ref{fig:trainsize} shows the AUC and the train
time depend on the log of the size of the training set.  In solid blue
lines, we see that the AUC improves almost linearly with the log of
the training size set.  In dotted red, we see that the train time
roughly scales logarithmically with the log of the train size, with an
exception of the data point corresponding to train size of 2000.

Each model was trained with the same HOG parameterization from the
grid search and the optimal regularization parameter for that subset
train data.  The train time of a given model generally increases for
increasing regularization parameter.  For the subset of train size
2000, the optimal regularization parameter happened to be
$C_{LR}=100$, whereas it was $C_{LR}=500$ for the subset of trainsize
1000, and $C_{LR}=1000$ for the subset of trainsize 4000.

% Scores vs. regularization on larger train/test set with comparable
% grid search in parameterization
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                                        
\begin{figure}[t]\label{fig:trainsizeLSST10}
\begin{center}
\includegraphics[width=\columnwidth]{figures/trainsize/LSST10_with_nLSST10.pdf}
\caption{Solid (dotted) blue line: AUC for models with varying size of
  the training set for LSST10 (nLSST10).  Solid (dotted) red line:
  Optimal regularization parameter for training on the train size for
  LSST10 (nLSST10). The improvement of AUC scales linearly with the
  log of the training set size, but train time roughly scales
  logarithmically.  Note, LSST10 requires more model complexity to
  exceed the performance of nLSST10.}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

%-------------------------------------------------%
\subsection{Machine Learning Algorithm: Logistic Regression}\label{sec:LR}
%-------------------------------------------------%
The problem of detecting gravitational lenses in images falls under
the general category of {\em{classification}} in machine learning. In
general the task is to find a function that assigns data points $x$ to
one of two or more classes, denoted by the class label $y$.  This is
equivalent to specifying a decision boundary, or decision boundaries
between the classes in the space of the data points.  (Compare this to
{\em{regression}} in which the task is to find a function $y=f(x)$,
where $y$ is a continuous, rather than discrete, variable.) In our
case, we have two classes: lens and non-lens containing images, and
the data points $x$ are the HOG feature vectors transformed from the
training set data.  In this paper we use the Logisitic Regression (LR)
algorithm, for which the decision boundary is a hyperplane .  (The
equivalent in the regression setting would be linear regression.) In
LR we determine the optimal hyperplane by minimizing the objective
function
\begin{equation}
\label{eqn:log_reg_objective}
L(A,b) = \sum_i \log \left[1 + \exp \left(-y_i (A \cdot x_i + b) \right) \right]
\end{equation}
where $x_i$ is a data point (HOG feature vector), $y_i$ is the
known label for that data point ($1$ for a lens containing
image, $-1$ for a non-lens containing), and $A$ and $b$ are
the parameters of the hyperplane. Eq. \ref{eqn:log_reg_objective}
is to be minimized with respect to $A$ and $b$.
Other more complicated machine learning algorithms exist which
do not necessarily produce a linear decision boundary, such as
Support Vector Machines (SVM), Random Forests, and Neural Networks.
training set data \citep{hastie_09}.

The HOG feature vectors in this paper can be very high-dimensional.
When dealing with high-dimensional data, where the number of
dimensions becomes comparable to the number of data points,
overfitting can become an issue. (An example of an extreme case
of overfitting would be if you fit a degree $n$ polynomial to $n$
points.  The polynomial would simply wiggle so that it goes
through every point, and would have no predictive power if you
tried to interpolate or extrapolate.) In machine learning,
overfitting is avoided using different {\em{regularization}}
techniques. A common choice for logistic regression is to
add a penalty term to Equation~\ref{eqn:log_reg_objective}
\begin{equation}\label{eqn:log_reg_regularized}
L_\text{Reg}(A,b,C_\text{LogReg}) = L(A,b) + \frac{1}{2C_\text{LogReg}} \|A\|
\end{equation}
where the norm $\|A\|$ is typically taken as either the
sum of squares of the coefficients ($L_2$ norm) or the sum
of absolute values of the coefficients ($L_1$ norm).  In this
paper we use the former.

The amount of regularization is controlled by the parameter
$C_\text{LogReg}$: larger values of $C_\text{LogReg}$ correspond to
increasing model complexity.  If $C_\text{LogReg}$ is too large then
the model will overfit, and if it is too small the model will
underfit.  To determine whether a model is overfit or underfit, the
model is trained, (i.e. Eq. \ref{eqn:log_reg_regularized} is
minimized), on a subset of the data called the {\em{training set}} and
its performance (goodness of fit) is evaluated both on the the
training set and a set of hold-out data called the {\em{test set}}
that was not used in constructing the model.
Figure~\ref{fig:regularization} shows the performance of a model for
selected HOG parameters as a function of the regularization parameter,
$C_\text{LogReg}$, for both the training and the test set.  (The
performance can be measured by the accuracy, i.e. percent of images
correctly classified, or by some other metric, such as the area under
the Receiver Operating Characteristic curve (see
Section~\ref{sec:ROC}).)

@@

% Logistic regression and regularization
As described in Section~\ref{sec:LR}, LR trains a model with
complexity determined by the regularization parameter coefficient,
$C_\text{LogReg}$.  Larger values of $C_\text{LogReg}$ are less
regularized and allow for increased model complexity.  The highest
values of $C_\text{LogReg}$ will better describe features in the
training set.  However, an overly complex model will overfit the
training set at the expense of its performance on any independent test
set.  The regularization parameter ultimately defines the model
performance, and we must perform a parameter search to identify the
optimal value for $C_\text{LogReg}$.

%  Our running over C - score and AUC as characterizations
Figure~\ref{fig:regularization} shows the model performance as a
function of regularization parameter for each data set HST, LSST, and
LSST10.  The solid and dotted blue lines respectively correspond to
the model performance on the test and training set, with the AUC as a
metric for performance.  In red, we show the train time as a function
of $C_{LogReg}$.

% Training behavior with C
As expected, the training set increases and asymptotes with
$C_\text{LogReg}$.  With increasing model complexity, the model better
fits the training data set.  This is analogous to fitting a seventh
order polynomial to seven data points, where the fitting function will
go through every point but will not likely predict additional points.
With increasing model complexity, we are able to better able to
capture features that are generally characteristic of strong lensing
systems with arcs. However, past a certain $C_\text{LogReg}$, the
model performance on the test set decreases or asymptotes, as it has
overfit the training set.  We use the scaling of AUC with
$C_\text{LogReg}$ when training 8,000 out of our full 10,000 training
data set to determine the best value for $C_\text{LogReg}$.  However,
the optimal parameter is also dependent on the size of the training
set (see Section~\ref{sec:trainsetsize}), so this choice is not
generalizable.

% Train time with Creg
For fixed train size, the log of the train time roughly scales
linearly with the log of $C_\text{LogReg}$.  Since lower values of
$C_\text{LogReg}$ correspond to a more regularized model, there is a
smaller volume in hyperparameter space to search for the best fit
coefficients.  The solution, on average, will converge more quickly,
for more regularized models.  The scaling is not purely monotonic
because the fitting still has some randomness associated with the path
it takes to convergence.

% Scores vs. regularization on larger train/test set with comparable
% grid search in parameterization
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                                        
\begin{figure*}[t]
\begin{center}
\includegraphics[width=.66\columnwidth]{figures/Creg/HST.pdf}
\includegraphics[width=.66\columnwidth]{figures/Creg/LSST1.pdf}
\includegraphics[width=.66\columnwidth]{figures/Creg/LSST10.pdf}
\caption{AUC of the model with varying LR regularization coefficient
  parameter, $C_{reg}$, used when training the model classifier.  We
  use a subset of the the 10,000 training images to search over the LR
  $C_{reg}$ parameter, training on 8,000 and testing on 1,000.  Each
  panel corresponds to a different mock observation.  From left to
  right: HST, LSST for one year, and LSST for 10 years.  The solid
  blue lines correspond to the AUC of the test set, and the dotted
  blue lines to the AUC of the training set.  To avoid overfitting, we
  choose the smallest parameter for which the AUC of the test set is
  maximal: 5000, 10, and 5000, respectively.  In thin red solid lines,
  we show the train time of the model, which roughly increases in a
  log-log scaling with logistic regression coefficient
  parameter.}\label{fig:regularization}
\end{center}
\end{figure*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%-------------------------------------------------%
\subsection{Effects of Dataset Size on Model Performance}\label{sec:datasize}
%-------------------------------------------------% 
%-------------------------------------------------% 
\subsubsection{Effects of Training Set Size}\label{sec:trainsetsize}
%-------------------------------------------------% 
In this section, we show the effects of training set size on model
performance.  Figure~\ref{fig:trainsizeLSST10} shows how the AUC and
the logistic regression coefficient depend on the log of the size of
the training set.  In solid blue lines, we see that the AUC improves
almost linearly with the log of the training size set.  In dotted red,
we see that the train time roughly scales logarithmically with the log
of the train size, with an exception of the data point corresponding
to train size of 2000.

Each model was trained with the same HOG parameterization from the
grid search and the optimal regularization parameter for that subset
train data.  The train time of a given model generally increases for
increasing regularization parameter.  For the subset of train size
2000, the optimal regularization parameter happened to be
$C_{LogReg}=100$, whereas it was $C_{LogReg}=500$ for the subset of trainsize
1000, and $C_{LogReg}=1000$ for the subset of trainsize 4000.

% Scores vs. regularization on larger train/test set with comparable
% grid search in parameterization
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                                        
\begin{figure}[t]
\begin{center}
\includegraphics[width=\columnwidth]{figures/trainsize/LSST10_with_nLSST10.pdf}
\caption{Solid (dotted) blue line: AUC for models with varying size of
  the training set for LSST10 (nLSST10).  Solid (dotted) red line:
  Optimal regularization parameter for training on the train size for
  LSST10 (nLSST10). The improvement of AUC scales linearly with the
  log of the training set size, but train time roughly scales
  logarithmically.  Note, LSST10 requires more model complexity to
  exceed the performance of nLSST10.}\label{fig:trainsizeLSST10}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
